---
title: "hw02_gap_dplyr"
author: "Avril Metcalfe-Roach"
date: "18 September 2019"
output: github_document
---

```{r libraries, warning=FALSE, include=FALSE}
library(dplyr)
library(tidyverse)
library(gapminder)
library(datasets) # For Q3
```

# Exercise 1: Dplyr package

## Data Initialization
```{r as_tibble}
raw_data <- gapminder %>% 
  as_tibble()
```


## 1.1 - Subset gapminder to 3 countries, 1970s.
### Countries: Canada, India, Italy
```{r 1.1}
filtered <- raw_data %>% 
  filter(country %in% c("Canada","India","Italy"), 
         year %in% c(1970:1979))
```

## 1.2 - Select country, gdpPercap using %>%
```{r 1.2}
country_gdp <- filtered %>% 
  select(country,gdpPercap)
country_gdp
```

## 1.3 - Countries with drops in life expectancy
```{r 1.3}
exp_list <- gapminder$lifeExp
change <- diff(exp_list,lag=1,differences=1)
# Add NA value to beginning of change vector:
change_2 <- append(change,NA,after=0)
# Create new tibble with delta life expectancy as a column:
gapminder_lifeExp <- gapminder
gapminder_lifeExp$delta <- change_2

# NOTE: INCLUDES APPARENT REDUCTIONS IN LIFE EXP DUE TO SWITCHING COUNTRIES!!!
gapminder_redExp <- gapminder_lifeExp %>% 
  filter(delta < 0)
```

## 1.4 - Gapminder: max GDP per capita per country
*Original question (using group_by())*
```{r 1.4}
# Create new column that lists the max GDP per country
gap_max_gdp <- gapminder %>% 
  group_by(country) %>% 
  mutate(max_gdp = max(gdpPercap)) %>% 
  ungroup()

# Filters gapminder to only show max GDP; removes the redundant 'max GDP' column
max_per_country <- gap_max_gdp %>% 
  filter(gdpPercap == max_gdp) %>% 
  subset(select = -max_gdp)
```

## 1.5 - Canadian Life Expectancy vs GDP
```{r 1.5}
# Select data
canadians <- gapminder %>% 
  filter(country=="Canada") %>% 
  select(lifeExp,gdpPercap)

# Plot in ggplot
ggplot(canadians, aes(gdpPercap,lifeExp)) +
  geom_point(alpha=0.5, colour = "red") +
  scale_x_log10("GDP per capita ($)") +
  ylab("Life Expectancy (years)")
```

# Exercise 2: Explore individual variables with dplyr

## Categorical variable: continent

### Possible range of continent
- Assuming we're not creating any new continents, this variable is inherently limited to the seven continents.
    - *Note*: North & South America are grouped into 'Americas'
- Possibilities: c(**Asia, Americas, Europe, Africa, Oceania**, Antarctica)
    - *Note*: Antarctica has no entries in gapminder, as it is a research base.

### Spread of values
Box and Whisker summary of data:
```{r con_values}
con_only <- raw_data %>% #gapminder as tibble
  select(continent)
con_sum <- count(con_only, continent) %>%  # dplyr: table of counts per continent
  as_tibble()
print(summary(con_sum))
boxplot(con_sum$n, 
        ylab="Number of data entries",
        xlab = "Continents")
```

The number of datapoints for each populated continent (e.g. Antarctica not included) **ranged from 24 to 624**. 
The **mean and median were 341 and 360** respectively, with 50% of the data falling between 300 and 396 entries.

Visual representation of distribution:
```{r con_graphs}
bar_plot <- ggplot(con_sum, aes(continent,n)) +
  geom_col(colour="black",fill="blue") +
  geom_text(aes(label=n), vjust=-0.25) +
  ylab("# of Entries") +
  xlab("Continent")

pie_plot <- ggplot(con_sum, aes(x='',y=n,fill=continent)) +
  geom_bar(width=1, stat = "identity") +
  coord_polar(theta="y")

require(gridExtra)
gridExtra::grid.arrange(bar_plot,pie_plot,ncol=2)
```

Generally, Oceania is very underrepresented, comprising just 24 out of 1704 entries. Conversely, African data was included at twice the rate of the average at 624 entries. The other three continents are relatively evenly represented.


## Quantitative variable: pop (population)

### Range of pop
The value of pop must be a Natural (>0) number. No strict upper limit is specified, but should logically be approximately 1.4 billion (the population of China).

```{r pop_range, warning=FALSE}
pop_only <- raw_data %>% # gapminder as tibble
  select(pop)
print(summary(pop_only))
boxplot(pop_only$pop, 
        ylab="Population",
        xlab = "")
```

As demonstrated by the boxplot, the vast majority of the data (all data within the whiskers/confidence interval) comprise a tiny fraction of the possible range of population values. 
50% of the data decribes a population between 2.8-19.6 million, with the median population being 7 million. The average is much higher at 29.6 million as the high-population outliers are skewing the data. The minimum and maximum populations are 60 000 and 1.32 billion respectively.

There are 12 entries for each country, as they were sampled at every time point. We can divide the data by year to see how the average populations change over time:

```{r pop_over_time}
# Linear Plot
pop_time <- raw_data %>% # raw_data is gapminder in tibble format 
  select(year, pop)
pop_time$year <- as.factor(pop_time$year)
pop_time_plot <- ggplot(pop_time, aes(year, pop)) +
  geom_boxplot() +
  xlab("Year") +
  ylab("Population") +
  ggtitle("Linear")

# Log Transformed
log_plot <- pop_time_plot + 
  scale_y_log10() +
  ggtitle("Logarithmic")

# Side by Side Output
require(gridExtra)
gridExtra::grid.arrange(pop_time_plot,log_plot,nrow=2)
```
 
The above graph makes it easier to see that there are only a couple of countries that have populations significantly outside of the statistical range. 
In the linear plot, the significant population size and fast growth of China and India in particular make the population growth of the rest of the world less apparent. 
By transforming the y axis to a log 10 scale, all of the outliers can be captured and the general trends become apparent: the IQR (middle 50%) of the data moves up the y axis, showing exponential population growth.

# Exercise 3: Plot Exploration
## Scatterplot of [CO2]ambient vs [CO2]uptake

Dataset: __CO2 - *Carbon Dioxide Uptake in Grass Plants*__
Here are the main parameters of the CO2 dataset:
```{r CO2_summary}
summary(CO2)
```
Plant is the type of plant, Type is the location of the plant, conc is the ambient CO2 concentration, and uptake is the CO2 absorbed by the plant.

```{r scatterplot}
scatter <- CO2 %>% 
  mutate(Group = (if_else
                  (Type=="Quebec",
                    if_else
                    (Treatment=="nonchilled","Qnc","Qc"),
                     if_else(Treatment=="nonchilled","Mnc","Mc"))))
ggplot(scatter, aes(conc,uptake)) +
  geom_point(aes(colour=(factor(Group))))
```

The above graph shows the change in CO2 uptake as a function of CO2 ambient concentration. In the legend, M/Q denote location (Mississippi vs Quebec) and c/nc denote treatment (chilled/not chilled).

From the plot, it can be seen that Quebec plants are more efficient at carbon fixation than Mississippi plants, and chilling the plants reduces their efficiency.

## Graph #2
Dataset: __esoph - *Smoking, Alcohol and (O)esophageal Cancer*__
Here are the main parameters of the CO2 dataset:
```{r esoph_summary}
esoph_cancer <- esoph %>% 
  rename("Alcohol Intake"=alcgp) %>% 
  rename("Tobacco Intake"=tobgp)
summary(esoph_cancer)
```
*Note: Ncases and ncontrols are not relevant here, as they supply weightings to the alcohol and tobacco groups.* 

We will be looking at the relationship between alcohol consumption and cancer rates over time.
```{r esoph_rates}

p<-esoph_cancer %>% 
  select(-`Tobacco Intake`)
p2 <- aggregate(cbind(p$ncases,p$ncontrols),
                by=list(esoph_cancer$agegp, esoph_cancer$`Alcohol Intake`), 
                FUN=sum)

esoph_format <- function(alcohol){
  p2 %>% 
    mutate(`Cancer Rate` = (V1/(V1+V2)*100)) %>% 
    select(Group.1,Group.2,`Cancer Rate`) %>% 
    filter(Group.2 == alcohol) %>% 
    return()
}

esoph_39 <- esoph_format("0-39g/day")
esoph_79 <- esoph_format("40-79")
esoph_119 <- esoph_format("80-119")
esoph_120 <- esoph_format("120+")

esoph_pie <- function(df){
  p3 <- ggplot(df, aes(x='',y='',fill=`Cancer Rate`)) +
      geom_bar(width=1,stat="identity") +
      coord_polar(theta="y")
      return(p3)
}
esoph_pie(esoph_120)
```

