---
title: "hw02_gap_dplyr"
author: "Avril Metcalfe-Roach"
date: "18 September 2019"
output: github_document
---

```{r libraries, warning=FALSE, include=FALSE}
library(dplyr)
library(tidyverse)
library(gapminder)
```

# Exercise 1: Dplyr package

## Data Initialization
```{r as_tibble}
raw_data <- gapminder %>% 
  as_tibble()
```


## 1.1 - Subset gapminder to 3 countries, 1970s.
### Countries: Canada, India, Italy
```{r 1.1}
filtered <- raw_data %>% 
  filter(country %in% c("Canada","India","Italy"), 
         year %in% c(1970:1979))
```

## 1.2 - Select country, gdpPercap using %>%
```{r 1.2}
country_gdp <- filtered %>% 
  select(country,gdpPercap)
country_gdp
```

## 1.3 - Countries with drops in life expectancy
```{r 1.3}
exp_list <- gapminder$lifeExp
change <- diff(exp_list,lag=1,differences=1)
# Add NA value to beginning of change vector:
change_2 <- append(change,NA,after=0)
# Create new tibble with delta life expectancy as a column:
gapminder_lifeExp <- gapminder
gapminder_lifeExp$delta <- change_2

# NOTE: INCLUDES APPARENT REDUCTIONS IN LIFE EXP DUE TO SWITCHING COUNTRIES!!!
gapminder_redExp <- gapminder_lifeExp %>% 
  filter(delta < 0)
```

## 1.4 - Gapminder: max GDP per capita per country
```{r 1.4}
# Create new column that lists the max GDP per country
gap_max_gdp <- gapminder %>% 
  group_by(country) %>% 
  mutate(max_gdp = max(gdpPercap)) %>% 
  ungroup()

# Filters gapminder to only show max GDP; removes the redundant 'max GDP' column
max_per_country <- gap_max_gdp %>% 
  filter(gdpPercap == max_gdp) %>% 
  subset(select = -max_gdp)
```

## 1.5 - Canadian Life Expectancy vs GDP
```{r 1.5}
# Select data
canadians <- gapminder %>% 
  filter(country=="Canada") %>% 
  select(lifeExp,gdpPercap)

# Plot in ggplot
ggplot(canadians, aes(gdpPercap,lifeExp)) +
  geom_point(alpha=0.5, colour = "red") +
  scale_x_log10("GDP per capita ($)") +
  ylab("Life Expectancy (years)")
```

# Exercise 2: Explore individual variables with dplyr

## Categorical variable: continent

### Possible range of continent
- Assuming we're not creating any new continents, this variable is inherently limited to the seven continents.
    - *Note*: North & South America are grouped into 'Americas'
- Possibilities: c(**Asia, Americas, Europe, Africa, Oceania**, Antarctica)
    - *Note*: Antarctica has no entries in gapminder, as it is a research base.

### Spread of values
Box and Whisker summary of data:
```{r con_values}
con_only <- raw_data %>% #gapminder as tibble
  select(continent)
con_sum <- count(con_only, continent) %>%  # dplyr: table of counts per continent
  as_tibble()
print(summary(con_sum))
boxplot(con_sum$n, 
        ylab="Number of data entries",
        xlab = "Continents")
```

The number of datapoints for each populated continent (e.g. Antarctica not included) **ranged from 24 to 624**. 
The **mean and median were 341 and 360** respectively, with 50% of the data falling between 300 and 396 entries.

Visual representation of distribution:
```{r con_graphs}
bar_plot <- ggplot(con_sum, aes(continent,n)) +
  geom_col(colour="black",fill="blue") +
  geom_text(aes(label=n), vjust=-0.25) +
  ylab("# of Entries") +
  xlab("Continent")

pie_plot <- ggplot(con_sum, aes(x='',y=n,fill=continent)) +
  geom_bar(width=1, stat = "identity") +
  coord_polar(theta="y")

require(gridExtra)
gridExtra::grid.arrange(bar_plot,pie_plot,ncol=2)
```

Generally, Oceania is very underrepresented, comprising just 24 out of 1704 entries. Conversely, African data was included at twice the rate of the average at 624 entries. The other three continents are relatively evenly represented.


## Quantitative variable: pop (population)
