---
title: "Analysis 2: Task Option 5"
output: 
  html_document:
    keep_md: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include = FALSE, warning=FALSE}
library(gapminder)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(formattable)
library(scales)
```

### Task Option 5: Life Expectancy over Time, Per Continent

The gapminder dataset was used to identify how life expectancy has changed over time, on a per-continent basis.

```{r o5}
o5 <- gapminder %>% 
  group_by(continent,year) %>% 
  summarize(`Avg Life Exp` = mean(lifeExp),
            `Std Dev Life Exp` = sd(lifeExp)) %>% 
  rename("Continent"=continent, "Year" = year)
o5$`Avg Life Exp` <- round(o5$`Avg Life Exp`,1)
o5$`Std Dev Life Exp` <- round(o5$`Std Dev Life Exp`,1)
o5
```

This ugly table summarizes the life expectancy per continent, per year by including the Avg Life Expectancy (`mean(lifeExp)`) and the Std Deviation of the Life Expectancy (`sd(lifeExp)`). Values were rounded to the nearest 0.1 year.

The change in life expectancy per continent was then graphed in a line chart:

```{r o5_plot}
o5 %>% 
  ggplot(aes(Year, `Avg Life Exp`, 
             group = Continent, 
             colour = Continent)) +
  geom_line(size = 1) +
  geom_point() +
  geom_errorbar(aes(ymin = `Avg Life Exp` - `Std Dev Life Exp`,
                    ymax = `Avg Life Exp` + `Std Dev Life Exp`), 
                size = 0.75, position=position_dodge(0.05)) +
  ggtitle("Change in Life Expectancy Per Continent") +
  theme_classic()
```

Error bars were included to indicate the spread of the data. The life expectancy of all continents increases over time, and most continents experience a reduction in the spread of life expectancies, indicating less disparity between different countries. African life expectancies, however, become more variable over time.

It is also interesting to note that all continents except for Africa seem to have converging life expectancies.

### Site Map

*[Assignment Repository](https://armetcal.github.io/html_hw)*

[Assignment 3 Home Page](https://armetcal.github.io/html_hw/hw03)

[Analysis 1](https://armetcal.github.io/html_hw/hw03/hw03.html)

**[Analysis 2 (Current Page)](https://armetcal.github.io/html_hw/hw03/hw03_2.html)**

[Analysis 3](https://armetcal.github.io/html_hw/hw03/hw03_3.html)


### References

Error bars added to plot based on guide posted by [sthda.com](http://www.sthda.com/english/wiki/ggplot2-error-bars-quick-start-guide-r-software-and-data-visualization).