---
title: 'Assignment #3: dplyr, ggplot2'
author: "Avril Metcalfe-Roach"
date: "9/26/2019"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include = FALSE, warning=FALSE}
library(gapminder)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(formattable)
library(scales)
```

# Assignment Overview
This assignment will use dplyr (via tidyverse) and ggplot2 pacakges to perform three analyses of the gapminder dataset, and the results will be compiled into a published html webpage.

Each analysis will produce a tibble (dplyr), accompanying plot (ggplot2), and a written explanation of the data. *Note: tibbles are reformatted as kables for aesthetic purposes using `knitr::kable()`.*

# Task Option 2: Max/min GDP per capita, per continent
The maximum and minimum recorded GDP per capita was extracted from the gapminder dataset for each continent, along with the accompanying country.

```{r option_2}
o2 <- gapminder %>% 
        group_by(continent) %>% 
          summarize(`Max GDP` = max(gdpPercap),
            `Max Country` = country[gdpPercap==`Max GDP`],
            `Max Year` = year[gdpPercap==`Max GDP`],
            `Min GDP` = min(gdpPercap),
            `Min Country` = country[gdpPercap==`Min GDP`],
            `Min Year` = year[gdpPercap==`Min GDP`])
o2$`Max GDP` <- formattable::currency(o2$`Max GDP`, digits=2L)
o2$`Min GDP` <- formattable::currency(o2$`Min GDP`, digits=2L)
knitr::kable(o2,format="html") %>% 
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>% 
  column_spec(1, bold = TRUE) %>% 
  column_spec(2, color = "green") %>% 
  column_spec(5, color = "red")
```
 <!--Bootstrap styling from kableExtra documentation:    https://haozhu233.github.io/kableExtra/awesome_table_in_html.html-->
 <!--The currency package:
 https://stackoverflow.com/questions/47664055/r-displaying-numbers-as-currency-in-knitrkable-->
 
 The table above demonstrates that there is a striking range of GDPs per capita across each continent, with differences of up to ~300x between the richest and poorest countries. It should be noted that some of the difference is attributable to inflation (e.g. for Europe and Oceania, the lowest/highest GDPs are found in the first/last years).
 
```{r o2_slope}
# Make plot:
ggplot(o2) +
  geom_segment(aes(x=1,xend=2,y=`Max GDP`, yend=`Min GDP`, col=continent),
               size=1, show.legend=TRUE) +
  geom_vline(xintercept = 1, size = 0.5) +
  geom_vline(xintercept = 2, size=0.5) +
  labs(x='',y="GDP per Capita") +
  xlim(1,2) +
  ylim(0,(1.15*(max(o2$`Max GDP`)))) +

  # Add labels:
  geom_text(label = "Max GDP", x=1.16, y=1.1*max(o2$`Max GDP`), 
            hjust=1.2, size = 4) +
  geom_text(label = "Min GDP", x=1.87, y=1.1*max(o2$`Max GDP`), 
            hjust=-0.1, size = 4) +
  ggtitle("Highest and Lowest GDPs per Capita on Record (Per Continent)") +

# Format nicely!
  theme(panel.background = element_blank(),
          panel.grid=element_blank(),
          axis.ticks=element_blank(),
          axis.text.x=element_blank(),
          panel.border = element_blank(),
          legend.title = element_blank(),
          legend.position = "bottom",
          plot.margin = unit(c(1,1,1,1), "cm"))
```
<!--http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#Animated%20Bubble%20Plot-->

These data were then plotted into a slope graph, as seen above. By plotting the maximum and minimum GDPs on the left and right axes and connecting them by continent, two key factors can be easily analyzed: 

**1. Difference between highest and lowest GDPs of any continent** - Asia (green) has a particularly large difference between the two GDPs of interest, whereas all of the other continents have a relatively consistent spread of GDPs.

**2. Relative trends between continents** - The spread of highest GDPs is larger than that of the lowest GDPs. This can be rationalized; there is a natural lower limit for GDPs (zero), whereas there is no meaningful upper limit; wealth tends to increase exponentially.
